Refining from the BON High-level Specification Language to the VHDL-AMS Hardware Description Language

### 2009-08-26 papers/BON_VHDL/paper.tex
refined paper

\lstdefinelanguage{Bon} {
  morekeywords={require,ensure,feature,old,and,o,Q'slew,end,Result, Void},
  morekeywords={class_chart,class,delta,invariant,inherit,deferred,effective,NONE,make},
  morekeywords={explanation,query,command,constraint},
  morecomment=[l]{--}, 
  morecomment=[s]{/*}{*/}, 
  morestring=[b]", 
 literate=,
  showstringspaces=false
}

@@

\newcommand{\tablesize}{\footnotesize}
\newcommand{\eg}{e.g.,\xspace}
\newcommand{\ie}{i.e.,\xspace}
\newcommand{\etc}{etc.\xspace}
\newcommand{\myhref}[2]{\ifpdf\href{#1}{#2}\else\htmladdnormallinkfoot{#2}{#1}\fi}
%\newcommand{\myhref}[2]{\emph{#2}}
\newcommand{\todo}{\textbf{TODO:}}
\newcommand{\ST}{\emph{SenseTile}\xspace}
\newcommand{\STs}{\emph{SenseTiles}\xspace}
\newcommand{\STS}{\emph{SenseTile Simulator}\xspace}
\newcommand{\datastore}{\STs Scientific Datastore\xspace}
\newcommand{\computefarm}{The \STs Scientific Compute Farm\xspace}
\newcommand{\computefarmlong}{UCD CASL \STs Software and Data Compute Server Farm\xspace}
\newcommand{\sensorfarm}{The \STs\xspace}
\newcommand{\ent}{\emph{entity}\xspace}
\newcommand{\arch}{\emph{architecture}\xspace}
\newcommand{\archs}{\emph{architectures}\xspace}

\newcommand{\undef}{\emph{undefined}\xspace}
\newcommand{\arcoss}{\emph{across}\xspace}
\newcommand{\through}{\emph{through}\xspace}
%---------------------------------------------------------------------
% New commands, macros, \etc
%---------------------------------------------------------------------

@@

The \STs, as an embedded system, have several unusual built-in sensors 
and emitters, sensor nodes can be extended via a general-purpose USB bus, 
and \STs contain an FPGA that is dynamically updated on a per-experiment 
basis.  \STs are custom-designed in VHDL and VHDL-AMS.  Like many hardware-based systems,
 to write new software for or against the \ST architecture one either uses
the actual hardware or one runs a \STS.  But this simulator emulates a
piece of hardware, one would like to know if the simulator
\emph{actually} simulates the hardware.  We have developed an approach
through which we refine the VHDL specification up to the formal BON
specification language, and from this specification we generate an
executable, functional model-based specification in the JML language.
We have used this specification to implement a fully validated and
verified \ST simulator.

@@

%======================================================================
\section{Introduction}
The \STs, represents heterogeneous collection of digital, analog 
and mixed signal hardware components. Embedded system complexity 
requires the need to strive for formal descriptions each component 
separatelly in order to produce specification of the \STs simulator. 
The development of complex system requires sophisticated tool and methodology 
in order to be able to complete a correct design in time.  
The authors will illustrate in this paper how BON can represent a high 
level analogue mapping and will further illustrate a procedure for 
transformation from VHDL-AMS to BON. As an extremly simple example, 
we will give practical aspect of refinement from VHDL-AMS to BON,
applied on A/D convertor, in order to produce formal specification 
and software simulation. A comprehensive list of the mapping between BON 
and VHDL-AMS can be found in the appendix.
%=====================================================================

\section{VHDL-AMS}
IEEE VHDL-AMS 1076.1 is the industry standard mixed-signal 
high-level description language for electronic and multi-domain 
systems. VHDL-AMS, as an superset of VHDL, supports modeling and 
simulation of analog and mixed-signal systems. The main contribution 
of VHDL-AMS is ability to describe the behavior of complex system 
trought differential and algebraic equations (DAEs). For this purpose, 
VHDL-AMS introduces the following capabilities and therms:

\begin{itemize}
\item 
\emph{Terminals} represents phisical connection points or circuit nodes 
of a network connectivity. They are declared to be of various 
\emph{natures}, which represent different energy domains of a system.
Terminals hold no values; their only role is to facilitate the formation
of conservative equation sets that, in turn, constrain the associated 
\emph{branch quatities}. 
\item
The \emph{simultaneous statements} creates one or more characteristic 
equations that are solved by the dedicated simulation kernel a.k.a.
"analog solver", to model continuously valued behaviour.  
\item
The \emph{quantities} denote a waveform or a time series of values.
The behaviour of the \emph{quantities} is given by the \emph{simultaneous
 statements}. They take values as a result of solving the set of simultaneous 
ordinary differential and algebraic equations (DAEs). To simplify, 
quantities can be understood as analog unknowns of (DAEs) that specify 
analog behavior. There are three kinds of quantity in VHDL-AMS: free, 
branch and source quantities. A \emph{free quantity} can be used in 
signal-flow modeling. A \emph{branch quantity} is used to model conservative energy
systems. A \emph{source quantity} is used for frequency and noise modeling.
\item
The \emph{break statement} is an construct which allows to 
initialize the DAE system at the beginning of a transient 
simulation and to mark discontinuities during such a transient 
simulation. A break statement allows restart of the analog kernel,
than overrides quantity values and executes concurrently with the analogue model. 
For any concurrent break statement there is an associated process.
\end{itemize}

\section{Choice of BON as the modeling tool}
%Brief description and motivation.
The Business Object Notation (BON) is an analysis and design notation 
for object-oriented systems  which emphasizes \emph{seamlessness}, 
\emph{reversibility},software \emph{contracting} and \emph{refinement}. 
Let us define these principles precisely and point out why the principles 
are important in VHDL-AMS simulation.

\subsection{Seamlessness and Reversibility}
\emph{Seamlessness} is the use of a continuous process 
throughout the software lifecycle.
\emph{Reversibility} is the support for both forward and 
backward development process: from analysis to design 
and implementation, and back.
In VHDL-AMS, modelling represents an continious top-down-bottom-up
process. Top-down approach serves for design while bottom-up
serves for verification. The possibility of seamlessly translating 
changes made during a certain development phase back into earlier phases,
facilitates a stable and predictable environment and ensures the various 
models consistent with the code. Keeping in sync various models, documetation
and specification from forward progress to backward and redesign, preclude 
that design and implementation diverge to an inconsistent state. 
\emph{Seamlessness} also plays important role in relating behaviours at 
differing levels of abstraction. For example, one may wish to relate the 
structural views of an A/D converter at the level of individual bits and 
at the level of busses. Therefore capturing similarities between the 
differing levels of abstractions far outweigh their inevitable differences, 
making it possible to obtain a continuous process that ensures a direct 
mapping between a formal specification and its software solution.

\subsection{Contracting and Refinement}
\emph{Contracting} is the definition, for each software element, 
of the associated semantic properties. BON use a full assertion mechanism, 
allowing analysts to specify not only the structure of a system but also its
semantics ( constraints and invariants of the expected results ). 
In VHDL-AMS \emph{simultaneous statement} represents analogue constraint of 
continuous-time waveform. In other words, analog kernel calculate quantities such 
that the constraints specified by the simultaneous statements are satisfied. 
We will agree that the constraint of waveform is nothing else than \emph{invariant}.
Another example of \emph{invariant} is conservation of energy law in which 
the summation of \arcoss quantities (\texttt{$\sum_{loop}$ across quatities =0}) 
around any loop equals zero and the summation of through 
quantities (\texttt{$\sum_{node}$ through quatities =0}) at any node equals zero. 
Moreover, any predefined (implicite) attribute in VHDL 
has defined pre(post) condition trought input(output) 
parameters and theirs constraints. Furthermore any subtype(subnature) 
represents restricted set of values from a base type(nature). 
The condition that determines which values are in the subtype(subnature)  
may be contracted by specifying characteristic predicate. 
For example, in typed $\lambda$ calculi, the VHDL-AMS subtype \lstinlinenc|bit| of 
\lstinlinenc|integer|, may be defined as \texttt{bit:=$\lambda$ n:nat.n < 2}. 
The corresponding informal and formal BON transformations are shown 
by the example below:
\input{bitbon}
Another advantage of BON is an easily-readable, English-based textual notation. 
The current IEEE standard VHDL-AMS language reference manual tries to define 
VHDL-AMS as well as possible in a descriptive way, explaining the 
semantics in English.\\
Therefore, on the basis of the above, BON is the obvious choice 
for the \ST simulator.
\section{Mapping between VHDL-AMS and BON}

\section{Refinement from VHDL-AMS to BON }

@@

\section {Acknowledgments}
The research upon which this paper is based was caried out with
sponsorship from Science Foundation Ireland via the 
\textbf{UCD CASL SenseTile System} grant.

>>>>

\lstdefinelanguage{Bon} {
  morekeywords={require,ensure,feature,old,and,o,Q'slew,end,Result, Void},
  morekeywords={class_chart,class,delta,invariant,inherit,deferred,effective,NONE,make},
  morekeywords={explanation,query,command,constraint,exists},
  morecomment=[l]{--}, 
  morecomment=[s]{/*}{*/}, 
  morestring=[b]", 
 literate=,
  showstringspaces=false
}

@@

\newcommand{\tablesize}{\footnotesize}
\newcommand{\eg}{e.g.,\xspace}
\newcommand{\ie}{i.e.,\xspace}
\newcommand{\etc}{etc.\xspace}
\newcommand{\myhref}[2]{\ifpdf\href{#1}{#2}\else\htmladdnormallinkfoot{#2}{#1}\fi}
%\newcommand{\myhref}[2]{\emph{#2}}
\newcommand{\todo}{\textbf{TODO:}}
\newcommand{\ST}{\emph{SenseTile}\xspace}
\newcommand{\STs}{\emph{SenseTiles}\xspace}
\newcommand{\STS}{\emph{SenseTile Simulator}\xspace}
\newcommand{\datastore}{\STs Scientific Datastore\xspace}
\newcommand{\computefarm}{The \STs Scientific Compute Farm\xspace}
\newcommand{\computefarmlong}{UCD CASL \STs Software and Data Compute Server Farm\xspace}
\newcommand{\sensorfarm}{The \STs\xspace}
\newcommand{\ent}{\emph{entity}\xspace}
\newcommand{\arch}{\emph{architecture}\xspace}
\newcommand{\archs}{\emph{architectures}\xspace}
\newcommand{\inv}{\emph{invariant}\xspace}
\newcommand{\undef}{\emph{undefined}\xspace}
\newcommand{\arcoss}{\emph{across}\xspace}
\newcommand{\through}{\emph{through}\xspace}
%---------------------------------------------------------------------
% New commands, macros, \etc
%---------------------------------------------------------------------

@@

The \STs, as an embedded system, have several unusual built-in sensors 
and emitters, sensor nodes can be extended via a general-purpose USB bus, 
and \STs contain an FPGA that is dynamically updated on a per-experiment 
basis.  \STs are custom-designed in VHDL-AMS.  Like many hardware-based systems,
 to write new software for or against the \ST architecture one either uses
the actual hardware or one runs a \STS.  But this simulator emulates a
piece of hardware, one would like to know if the simulator
\emph{actually} simulates the hardware.  We have developed an approach
through which we refine the VHDL specification up to the formal BON
specification language, and from this specification we generate an
executable, functional model-based specification in the JML language.
We have used this specification to implement a fully validated and
verified \ST simulator.

@@

%======================================================================
\section{Introduction}
The \ST sensor platform, represents heterogeneous collection of digital, analog 
and mixed signal hardware circuits. Embedded system complexity 
requires the need to strive for formal descriptions each component 
separatelly in order to produce specification of the \STS. 
Our aim is to be able to precisely describe the structures of circuits,
the behaviour of circuits and the formal link between structures and behaviours.
In other words, the behavioural specification of the \STS may be 
specified as a \inv on the signals at the circuit's ports and may be obtained 
by taking the conjunction of the \emph{invariants} that describe 
behaviours of the individual components separately. 
For this purpose we need  sophisticated tool and methodology 
in order to be able to achieve this goal.  
The authors will illustrate in this article how BON can represent a high 
level analogue mapping and will further illustrate a procedure for 
transformation from VHDL-AMS to BON.\\ 
This article provides two main contributions.
\begin{enumerate}
\item 
 It provides a solution how to formally emulate
 piece of hardware using BON-JML approach; up to date BON and JML
 has been used mostly in the pure object oriented software construction.    
\item
It relates hardware description code and formal calculus. 
This means that if there  is a code that is "correct by construction", 
then it will be derivable from informal BON specification, 
via the refinement calculus to formal BON specification 
and JML implementation.     
\end{enumerate}
As an extremly simple example, 
we will give practical aspect of refinement, applied on real capacitor component, 
in order to produce formal specification  and software simulation.\\  
A comprehensive list of the mapping between BON and VHDL-AMS can be 
found in the appendix.
%=====================================================================

\section{VHDL-AMS}
IEEE VHDL-AMS 1076.1 is the industry standard mixed-signal 
high-level description language for electronic and multi-domain 
systems. VHDL-AMS, as an superset of VHDL, supports modeling and 
simulation of analog and mixed-signal systems. The main contribution 
of VHDL-AMS is ability to describe the behaviour of complex system 
trought differential and algebraic equations (DAEs). For this purpose, 
VHDL-AMS introduces the following capabilities and therms:

\begin{itemize}
\item 
\emph{Terminals} represents phisical connection points or circuit nodes 
of a network connectivity. They are declared to be of various 
\emph{natures}, which represent different energy domains of a system.
Terminals hold no values; their only role is to facilitate the formation
of conservative equation sets that, in turn, constrain the associated 
\emph{branch quatities}. 
\item
The \emph{quantities} denote a waveform or a time series of values.
The behaviour of the \emph{quantities} is given by the \emph{simultaneous
 statements}. They take values as a result of solving the set of simultaneous 
ordinary differential and algebraic equations (DAEs). To simplify, 
quantities can be understood as analog unknowns of (DAEs) that specify 
analog behaviour. There are three kinds of quantity in VHDL-AMS: free, 
branch and source quantities. A \emph{free quantity} can be used in 
signal-flow modeling. A \emph{branch quantity} is used to model conservative energy
systems. A \emph{source quantity} is used for frequency and noise modeling.
\item
The \emph{simultaneous statements} creates one or more characteristic 
equations that are solved by the dedicated simulation kernel a.k.a.
"analog solver", to model continuously valued behaviour.  
\item 
The \emph{procedural statements} or \emph{procedural} express analog behaviour 
using sequential statements. Actually procedural describes analog behaviour 
by computing new values for quantities from the current value of quantities.
\item
The \emph{break statement} is an construct which allows to 
initialize the DAE system at the beginning of a transient 
simulation and to mark discontinuities during such a transient 
simulation. A break statement allows restart of the analog kernel,
than overrides quantity values and executes concurrently with the analogue model. 
For any concurrent break statement there is an associated process.
\end{itemize}

\section{Choice of BON as the modeling tool}
%Brief description and motivation.
The Business Object Notation (BON) is an analysis and design notation 
for object-oriented systems  which emphasizes \emph{seamlessness},\xspace 
\emph{reversibility} and software \emph{contracting}. 
Let us define these principles precisely and point out why the principles 
are important in VHDL-AMS simulation.

\subsection{Seamlessness and Reversibility}
\emph{Seamlessness} is the use of a continuous process 
throughout the software lifecycle.
\emph{Reversibility} is the support for both forward and 
backward development process: from analysis to design 
and implementation, and back.
In VHDL-AMS, modelling represents an continious top-down-bottom-up
process. Top-down approach serves for design while bottom-up
serves for verification. The possibility of seamlessly translating 
changes made during a certain development phase back into earlier phases,
facilitates a stable and predictable environment and ensures the various 
models consistent with the code. Keeping in sync various models, documetation
and specification from forward progress to backward and redesign, preclude 
that design and implementation diverge to an inconsistent state. 
\emph{Seamlessness} also plays an important role in relating behaviours at
different levels of abstraction. For example, one may wish to analyze
a "real capacitor" at the level of individual components (a resistor and ideal capacitor)
and their behaviours, or at the more abstract level through general behaviour (\inv) and
using existential quantification (\texttt{exists}) to encapsulate all internal signals. 
Therefore capturing similarities between the differing levels of abstractions 
far outweigh their inevitable differences, making it possible to obtain a 
continuous process that ensures a direct mapping between a formal specification 
and its software solution.

\subsection{Contracting}
\emph{Contracting} is the definition, for each software element, 
of the associated semantic properties. BON use a full assertion mechanism, 
allowing analysts to specify not only the structure of a system but also its
semantics ( constraints and invariants of the expected results ). 
In VHDL-AMS \emph{simultaneous statement} represents analogue constraint of 
continuous-time waveform. In other words, analog kernel calculate quantities such 
that the constraints specified by the simultaneous statements are satisfied. 
We will agree that the constraint of waveform is nothing more than \inv.
Another example of \inv is conservation of energy law in which 
the summation of \arcoss quantities (\texttt{$\sum_{loop}$ across quatities =0}) 
around any loop equals zero and the summation of through 
quantities (\texttt{$\sum_{node}$ through quatities =0}) at any node 
(\texttt{terminal}) equals zero. Moreover, any predefined (implicite) attribute in VHDL-AMS  
has defined pre (post) condition trought input (output) 
parameters and theirs constraints. Furthermore any subtype (subnature) 
represents restricted set of values from a base type (nature). 
The condition that determines which values are in the subtype (subnature)  
may be contracted by specifying characteristic predicate which represents \inv.\\ 
Another advantage of BON is an easily-readable, English-based textual notation. 
The current IEEE standard VHDL-AMS language reference manual tries to define 
VHDL-AMS as well as possible in a descriptive way, explaining the 
semantics in English.\\
Therefore, on the basis of the above, BON is the obvious choice 
for the \STS.
Throughput this article we will make this more clearly.  
\section{Mapping between VHDL-AMS and BON}
The main theme of this article is the mapping between VHDL-AMS and BON.
\section{Refinement from VHDL-AMS to BON }

@@

\section {Acknowledgments}
The research upon which this article is based was caried out with
sponsorship from Science Foundation Ireland via the 
\textbf{UCD CASL SenseTile System} grant.
